Untitled

### 2022-03-18 Papers/Polymethylation_scores/01_titlepage.Rmd
updated results post datafreeze update
colter revisions
post submission revisions

---
output:
  html_document: default
  word_document: default
bibliography: Methylation_SmokingBib.bib
always_allow_html: true
---

@@

**Title:** Polymethylation Scores for Prenatal Maternal Smoke Exposure Persist Until Age 15 and Are Detected in Saliva

**Authors:** Freida A. Blostein$^1$, Jonah Fisher$^2$,  John Dou$^1$, Lisa Schenper$^3$, Erin Ware$^2$ Daniel A. Notterman$^3$, Colter Mitchell$^2$, Kelly M. Bakulski$^1$`

@@

>>>>

---
output:
  pdf_document: default
  word_document: default
  html_document: default
bibliography: Methylation_SmokingBib.bib
always_allow_html: yes
---

@@

**Title:** Polymethylation Scores for Prenatal Maternal Smoke Exposure Persist Until Age 15 and Are Detected in Saliva in the Fragile Families and Child Wellbeing Cohort

**Authors:** Freida A. Blostein$^1$, Jonah Fisher$^2$,  John Dou$^1$, Lisa Schenper$^3$, Erin Ware$^2$ Daniel A. Notterman$^{3*}$, Colter Mitchell$^{2*}$, Kelly M. Bakulski$^{1*+}$

@@

*Co-senior authors

$^+$Corresponding author

### 2022-03-18 Papers/Polymethylation_scores/02_abstract.Rmd
post submission revisions

@@

**Background**: Prenatal maternal smoking has negative implications for child health.  DNA methylation signatures  can function as a biomarker of prenatal maternal smoking. However little work has assessed how DNA methylation signatures of prenatal maternal smoking vary across ages, ancestry groups, or tissues. In the Fragile Families and Child Wellbeing study, we tested whether prenatal maternal smoking was associated with salivary polymethylation scores. We assessed the consistency of associations at ages 9 and 15, their portability across  participants from African, European, and Hispanic genetic ancestries and the accuracy of exposure classification using Area Under the Curve (AUC) from receiver operating curve analyses.

**Results**: We created saliva polymethylation scores using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cordblood. In the full sample at age 9 (n=756), prenatal maternal smoke exposure was associated with a 0.52 (95%CI: 0.36, 0.67) standard deviation higher polymethylation score. The direction and magnitude of the association was consistent across when stratified genetic ancestries. In the full sample at age 15 (n=749), prenatal maternal smoke exposure was associated with a 0.46 (95%CI: 0.3, 0.62) standard deviation higher polymetylation score, and the association was attenuated among the European and Hispanic genetic ancestry samples. The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (AUC age 9=0.77, *P* value<0.001, age 15=0.77, *P* value<0.001), and more accurate than using a single *a priori* site in the AHRR gene (cg05575921 AUC=0.74, *P* value=0.03; age 15=0.73, *P* value=0.01).

**Conclusions**: Prenatal maternal smoking was associated with DNA methylation signatures in saliva samples, a clinically practical tissue. Polymethylation scores for prenatal maternal smoking were portable across genetic ancestries and more accurate than individual DNA methylation sites. DNA polymethylation scores from saliva samples could serve as robust and practical clinical biomarkers of prenatal maternal smoke exposure.

>>>>

---
output:
  pdf_document: default
  html_document: default
---

@@

**Background**: Prenatal maternal smoking has negative implications for child health.  DNA methylation signatures can function as biomarkers of prenatal smoke exposure. However the robustness of prenatal maternal smoking DNA methylation signatures across child ages, genetic ancestry groups or tissues is not clear. Using coefficients from a meta-analysis of prenatal maternal smoke exposure and DNA methylation in newborn cord blood, we created polymethylation scores of saliva DNA methylation from children at ages 9 and 15 in the Fragile Families and Child Wellbeing study.  We tested whether prenatal maternal smoking was associated with salivary polymethylation scores for prenatal smoke exposure and assessed the consistency of associations by age and participant genetic ancestry group. We tested the accuracy of exposure classification using area under the curve (AUC) from receiver operating curve analyses and compared to findings from candidate DNA methylation sites.

**Results**: In the full sample at age 9 (n=`r table(modeldata$childteen)[1]`), prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 9' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score. The direction and magnitude of the association was consistent in European and African genetic ancestry samples. In the full sample at age 15 (n=`r table(modeldata$childteen)[2]`), prenatal maternal smoke exposure was associated with a `r coefCI(global_models, "childteen=='Age 15' & predictors=='base model' & str_detect(outcome, 'cell-type')", 'smkPreg_binaryYes', 2)` standard deviation higher polymethylation score, and the association was attenuated among the European and Admixed - Latin genetic ancestry samples. The polymethylation score was reasonably accurate  at classifying prenatal maternal smoke exposure (AUC age 9=0.77, age 15=0.77). Including the polymethylation score increased the AUC of base model covariates by `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, '6074'))%>%pull(6)` percentage points, while including a single candidate site in the *AHRR* gene did not (*P* val `r rct1%>%filter(childteen=='Age 9' & str_detect(modeltype, 'cg055'))%>%pull(7)%>%round(.,2)`).

**Conclusions**: Prenatal maternal smoking was associated with DNA methylation signatures in child saliva samples, a clinically practical tissue. Polymethylation scores for prenatal maternal smoking were portable across genetic ancestries and more accurate than individual DNA methylation sites. DNA polymethylation scores from saliva samples could serve as robust and practical clinical biomarkers of prenatal maternal smoke exposure.

### 2022-03-18 Papers/Polymethylation_scores/04_background.Rmd
post submission revisions

---
output:
  word_document: default
  pdf_document: default
  html_document: default
bibliography: Methylation_SmokingBib.bib
latex_engine: xelatex
---

# Background

In 2018, 11% of pregnant women in the United States aged 15-44 reported any past-month cigarette use [@SAMHSA2018]. Children exposed to cigarette smoking in utero are more likely to have low-birth weight, negative neurodevelopmental outcomes and asthma [@Banderali:2015vu]. Measuring prenatal maternal smoking behavior is challenging. Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua]. The gold-standard biomarker, serum cotinine levels, have a half-life of nine hours in pregnant women, and are rarely available outisde of birth cohorts[@Dempsey:2002wr]. Alternative biomarkers of prenatal maternal smoke exposure could improve research and clinical practice.

DNA methylation is a possible biomarker. Prenatal maternal smoking is associated with DNA methylation signatures in placental, cord blood, and peripheral blood samples [@Joubert:2016ue; @Everson:2021td; @Cardenas2019; @Richmond:2018tu; @Tehranifar:2018tb]. Cord blood and placental samples are rarely available outside of birth cohorts. While peripheral blood is a more realistic sample, saliva is easier to collect [] and increasingly used in large epidemiological cohorts []. Since DNA methylation drives cell differentiation and cell type proportions differ across tissue types, it is important to verify associations in new tissues. To our knowledge, no study has evaluated associations between maternal prenatal cigarette smoking and salivary DNA methylation.

Additionally, DNA methylation can vary by genetics [@Galanter2017; @Fraser2012; @Moen2013; @Rahmani2017]. Own-smoking DNA methylation signatures differ by genetic ancestry [@Barcelona2019; @Sun2013; @Philibert2013; @Philibert2012; @Liu2019]. Most work on prenatal maternal smoking and child DNA methylation has been conducted in European-ancestry cohorts [@Joubert:2016ue], with some notable exceptions [@Ladd-Acosta:2016aa; @Tehranifar:2018tb; @Zhang2018; @Gonseth2016; @Neophytou:2019ty]. Some CpG sites associated with prenatal maternal smoking in European-ancestry cohorts have been replicated in non-European-ancestry or multi-ethnic populations [@Tehranifar:2018tb; @Neophytou:2019ty]. Children of both Black and non-Black identifying mothers exhibited similar directions of association between prenatal maternal smoking and peripheral blood DNA methylation at 38 CpG sites [@Xu2021]. However, this analysis stratified by self-reported ethnicity rather than genetic ancestry. Very studies have evaluated the portability of DNA methylation signatures of prenatal maternal smoke exposure across genetic ancestry populations within a single cohort [@Tehranifar:2018tb; @Gonseth2016].

After association testing, DNA methylation-based biomarkers of prenatal maternal smoking can be assessed by evaluating their classification accuracy. Biomarkers range from single DNA methylation sites to multi-site summary measures, such as polymethylation scores [@Philibert2020 @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa]. In a polymethylation score, CpG sites are weighted by their effect size from a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a]. DNA polymethylation scores for prenatal maternal smoke exposure have performed well (area under the curves (AUCs)\>0.8) in newborns and young children [@Reese:2017ts; @Ladd-Acosta:2016aa] and fairly in middle-aged adults (AUC of 0.72 (95% confidence interval 0.69, 0.76)) [@Richmond:2018tu]. DNA methylation signatures of own smoking change post-cessation [@keshawarz2021]. Associations with prenatal maternal smoking change at some CpG sites as children age [@Richmond2015]. Yet few studies have assessed the longitudinal persistence of DNA-methylation biomarker accuracy.

DNA methylation differences hold promise as a biomarker for prenatal maternal smoking. We must understand how and if the signal varies across tissues sample, genetic ancestry, and age. In the Fragile Families and Child Wellbeing study, a diverse longitudinal birth cohort, we aimed to assess salivary DNA methylation biomarkers for prenatal maternal cigarette smoking. We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including polymethylation scores for smoke exposure, individual a priori CpG sites, global DNA methylation, and epigenetic clocks, at ages 9 and 15. We then performed ancestry- and age-stratified analyses to test the hypothesis that DNA methylation could serve as a portable and persistent biomarker of prenatal maternal smoking.

>>>>

---
output:
  word_document: default
  pdf_document: default
  html_document: default
bibliography: Methylation_SmokingBib.bib
latex_engine: xelatex
csl: springer-vancouver.csl
---

# Introduction

In 2018, 11% of pregnant people in the United States aged 15-44 reported any past-month cigarette use [@SAMHSA2018]. Children exposed to cigarette smoking *in utero* are more likely to have low-birth weight, negative neurodevelopmental outcomes, and asthma [@Banderali:2015vu]. Measuring prenatal maternal smoking behavior can be challenging. Maternal prenatal smoking is underreported due to stigma [@Kvalvik:2012tk; @Dietz:2011ua]. The gold-standard smoking biomarker, serum cotinine levels, have a half-life of nine hours in pregnant women, and are rarely available outside of birth cohort research [@Dempsey:2002wr]. Alternative biomarkers of prenatal maternal smoke exposure with a larger window of measurement could improve research and clinical practice.

DNA methylation is a possible biomarker. Prenatal maternal smoking is associated with offspring DNA methylation signatures at birth in placental and cord blood samples, and postnatally in peripheral blood samples [@Joubert:2016ue; @Everson:2021td; @Cardenas2019; @Richmond:2018tu; @Tehranifar:2018tb]. Another postnatal biospecimen option is saliva, which is easier to collect than blood and increasingly used in large epidemiological cohorts [@Langie2017]. Since DNA methylation drives cell differentiation and cell type proportions differ across tissue types, it is important to verify associations in new tissues [@Bakulski2016]. In addition, most studies of prenatal maternal smoking and childhood DNA methylation have been conducted in primarily European genetic ancestry populations [@Joubert:2016ue; @Ladd-Acosta:2016aa; @Tehranifar:2018tb; @Zhang2018; @Gonseth2016; @Neophytou:2019ty], and it will be important to test generalizability to other groups. To our knowledge, few studies have assessed multiple genetic ancestry groups in the same cohort and no study has evaluated associations between maternal pregnancy cigarette smoking and salivary DNA methylation in children.

Using the findings from large-scale association testing, in independent populations, DNA methylation-based biomarkers of prenatal maternal smoking can be developed and assessed by evaluating their classification accuracy. Biomarkers can range from single DNA methylation sites to summary measures across multiple sites, termed polymethylation scores [@Philibert2020 @Philibert2013; @Reese:2017ts; @Richmond:2018tu; @Ladd-Acosta:2016aa]. In a polymethylation score, DNA methylation sites are weighted by their effect size from a previous, independent sample and then summed to a single score [@Chen2020; @Bakulski2021a]. DNA polymethylation scores for prenatal maternal smoke exposure have performed well (area under the curves (AUCs)\>0.8) in cord blood samples from newborns and peripheral blood samples from young children [@Reese:2017ts; @Ladd-Acosta:2016aa] and fairly well in peripheral blood from middle-aged adults (AUC of 0.72 (95% confidence interval 0.69, 0.76)) [@Richmond:2018tu]. Associations with prenatal maternal smoking change at some DNA methylation sites as children age [@Richmond2015]. Yet few studies have assessed the longitudinal persistence of DNA-methylation biomarker accuracy.

DNA methylation differences hold promise as a biomarker for prenatal maternal smoke exposure. We must understand how the signal varies across tissue, genetic ancestry, and age. In the Fragile Families and Child Wellbeing Study, a diverse longitudinal birth cohort, we aimed to assess salivary DNA methylation biomarkers for prenatal maternal cigarette smoking. We tested associations between prenatal maternal smoke exposure and saliva DNA methylation, including polymethylation scores for smoke exposure, individual a priori DNA methylation sites, global DNA methylation, and epigenetic clocks, at ages 9 and 15. We then performed genetic ancestry- and age-stratified analyses to test the hypothesis that DNA methylation could serve as a portable and persistent biomarker of prenatal maternal smoking.
